\documentclass[11pt,letterpaper]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amscd,amsthm,mathtools}
\usepackage{authblk}
\usepackage{booktabs}
\usepackage[hidelinks]{hyperref}
\usepackage{microtype}
\usepackage{enumitem}

\numberwithin{equation}{section}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

\newcommand{\NN}{\mathbb{N}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\Xfin}{X_\infty^{\mathrm{fin}}}
\newcommand{\Xinf}{X_\infty}
\DeclareMathOperator{\supp}{supp}
\DeclareMathOperator{\Val}{Val}
\DeclareMathOperator{\NF}{NF}
\DeclareMathOperator{\Fold}{Fold}

\renewcommand\Authfont{\normalsize}
\renewcommand\Affilfont{\small}
\setlength{\affilsep}{0.4em}

\title{Finite-window rigidity in Fibonacci numeration}
\author[1]{Haobo Ma}
\author[2]{Wenlin Zhang\thanks{Corresponding author.}}
\affil[1]{AELF PTE LTD., \#14-02, Marina Bay Financial Centre Tower 1, 8 Marina Blvd, Singapore 018981, Singapore\\\texttt{auric@aelf.io}}
\affil[2]{National University of Singapore, Singapore\\\texttt{e1327962@u.nus.edu}}
\date{}

\begin{document}
\maketitle

\begin{abstract}
For each $m\ge 1$, let $\Fold_m:\{0,1\}^m\to X_m$ send a binary word to the
first $m$ digits of the Zeckendorf expansion of its Fibonacci value.  We show
that $\Fold_m$ is the normal-form map of a finite terminating confluent rewrite
system and that the family $(\Fold_m)_{m\ge 1}$ is compatible with restriction
in $m$.  For $m\ge 3$ we prove that the associated sliding block code $\Phi_m$
has the following exact reconstruction property: every block of $n\ge m$
consecutive output symbols has a unique binary preimage of length $n+m-1$.
Equivalently,
the block map $B_{m,n}:\{0,1\}^{n+m-1}\to\mathcal L_n(Y_m)$ is a bijection and
$\lvert\mathcal L_n(Y_m)\rvert=2^{n+m-1}$.
Equivalently, for $m\ge 3$ the residue vector modulo $F_{m+2}$ of the
overlapping local Fibonacci sums determines the entire binary block uniquely.

This block bijection gives local inverse rules with memory $r$ and anticipation
$m-1-r$ for every $0\le r\le m-1$, and the one-sided inverse memory $m-1$ is
best possible.  It also yields explicit formulas for Bernoulli measures pushed
forward from the full two-shift: if $\mu_{m,p}=(\Phi_m)_*\nu_p$, then
$\mu_{m,p}([a])=p^{|\omega(a)|_1}(1-p)^{n+m-1-|\omega(a)|_1}$ for
$a\in\mathcal L_n(Y_m)$, and
$H_{\mu_{m,p}}(Y_0^{n-1})=(n+m-1)h_2(p)$.  For Bernoulli input bits, this
arithmetic residue vector has an explicit finite-dimensional law and forms an
exact $m$-step Markov process.  We also show that the right Fischer cover of
$Y_m$ over the natural alphabet $X_m$ is the explicit suffix graph $G_m$, which
has $2^{m-1}$ states and $2^m$ edges.
\end{abstract}

\medskip
\noindent\textbf{Keywords.}
Zeckendorf expansion, Fibonacci numeration, finite-window rigidity, Bernoulli
transport, exact decoding, Markov order, finite-state normalization.

\medskip
\noindent\textbf{2020 Mathematics Subject Classification.}
Primary 11A63; Secondary 11B39, 68Q45, 37B10.


\input{sections/sec_introduction}

\input{sections/sec_normalization}

\input{sections/sec_rewrite}

\input{sections/sec_compatibility}

\input{sections/sec_overlapping}

\input{sections/sec_finite_state}

\input{sections/sec_rigidity}

\input{sections/sec_arithmetic}

\section*{Competing interests}

The author(s) declare none.

\section*{Declaration of generative AI and AI-assisted technologies in the manuscript preparation process}

During the preparation of this work the author(s) used OpenAI ChatGPT and
Anthropic Claude in order to assist with language polishing and editing of the
manuscript text.  After using these tools, the author(s) reviewed and edited
the content as needed and take(s) full responsibility for the content of the
published article.

\bibliographystyle{plain}
\bibliography{../../2026_golden_ratio_driven_scan_projection_generation_recursive_emergence/references}

\end{document}
